---
title: "World Development Indicators (2022)"
author: "Anny Huang"
date: "2026-02-25"
format:
  revealjs:
    theme: solarized
    transition: fade
    slide-number: true
    chalkboard: false
    code-fold: true
---

```{r}
#| include: false
library(tidyverse)
wdi<-read_csv("wdi.csv")
```

## Motivation

-   The World Development Indicators (WDI) provide a snapshot of economic and social conditions across countries.

-   This presentation summarizes patterns in GDP per capita, life expectancy, and unemployment for 2022.

## Data

-   One-year sample (2022) with 14 indicators

-   Unit of observation: country

-   Source: World Bank, *World Development Indicators* (WDI)

## Key Indicator 1: GDP per Capita
```{r}
#| echo: false
library(ggplot2)
ggplot(wdi, aes(x = gdp_per_capita)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of GDP per Capita (2022)",
    x = "GDP per Capita",
    y = "Number of Countries"
  )
```
Takeaway: GDP per capita is strongly right-skewed, with a small number of very high-income countries.

## Life Expectancy

```{r}
#| echo: false
ggplot(wdi, aes(x = life_expectancy)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Life Expectancy (2022)",
    x = "Life Expectancy (years)",
    y = "Number of Countries"
  )
```

Takeaway: Life expectancy is more concentrated than GDP per capita.

## GDP and Life Expectancy

```{r}
#| echo: false
ggplot(wdi, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  labs(
    title = "GDP per Capita vs Life Expectancy (2022)",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)"
  )
```

Takeaway: Higher GDP per capita is generally associated with higher life expectancy, with smaller gains at high income levels.

## Unemployment Rate

```{r}
#| echo: false
ggplot(wdi, aes(x = unemployment_rate)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Unemployment Rate (2022)",
    x = "Unemployment Rate (%)",
    y = "Number of Countries"
  )
```

Takeaway: Unemployment varies widely across countries.

## Conclusion

- GDP per capita shows large cross-country inequality.
- Life expectancy is more evenly distributed and positively related to income.
- Unemployment differs substantially across countries.

Next step: Extend this analysis across multiple years to study trends.

